NumPy---Array Computation for Python

### 2020-01-08 body.tex
Reorg based on discussion w/ Ralf and Stefan

% https://books.google.com/books?id=QQJODwAAQBAJ&pg=PA101&lpg=PA101&dq=%E2%80%9Cthe+secret+to+successful+hiring+is+this:+look+for+the+people+who+want+to+change+the+world.%E2%80%9D&source=bl&ots=D6l8JXfB20&sig=ACfU3U0x1X9Gh8k6MlNV0BqhBa_73Q0yfQ&hl=en&sa=X&ved=2ahUKEwj48pqg_oHgAhVohuAKHYlJAsc4ChDoATAAegQIChAB#v=onepage&q=%E2%80%9Cthe%20secret%20to%20successful%20hiring%20is%20this%3A%20look%20for%20the%20people%20who%20want%20to%20change%20the%20world.%E2%80%9D&f=false
% "The secret to successful hiring is this: look for the people who want to change the world.".  Marc Benioff.

@@

%At the time, commercial languages for interactive computing with numerical
%arrays were already well-established, including Matlab and IDL.
%They were developed and maintained by large groups of
%full-time professional programmers, supported by formal quality assurance,
%release managers, and writers of scientific documentation.
%In contrast, Numeric, and later NumPy, would continue to be developed almost
%entirely by scientists, including graduate students and faculty.
%%without specific funding and often to the detriment of their careers.
%% S: The "Back then" sentence sets up an expectation.  Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".
%NumPy had no dedicated funding, and development time
%was mostly contributed freely by students and researchers in their
%spare time.

%Table X1 about here: backgrounds of initial designers (mainly math, physics).

%Table X2 about here: backgrounds of major committers (mainly math, physics).

@@

Over the next 10 years, NumPy attracted more scientist developers.
The project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy.
The project has comprehensive, high-quality documentation,
integrated with the source
code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

@@

Sometimes, it is desirable to combine arrays of different dimensions.  One may, for example, want to add a scalar to a vector, or scale each column of an array by a specific scalar value.  This is done through {\em broadcasting}, where one of the arrays is virtually duplicated (i.e., without copying any data in memory), so that the shapes of the operands match.  Broadcasting is illustrated in \ref{fig:broadcasting}.

@@

\section*{Recent development}

%For many years, NumPy had no dedicated funding, and development time
%was mostly contributed freely by students and researchers in their
%spare time.  In-person meetings were sponsored off of individual's
%research grants, and while industry sponsored contributions were made,
%notably those by Mark Wiebe when he was sponsored by Travis Oliphant
%at Enthought, the package's rapid adoption in scientific
%computing took place without external investment.

@@

As Python's popularity for scientific computing, data science, and machine
learning increased, many new use cases arose.
At the same time, the massive number of scientific Python packages that
built on NumPy meant that it had an unusually high need for stability.
So to guide our development we formalized the feature proposal process, and
constructed a development roadmap with extensive input and feedback from the
community.

In 2018, we hired two (and later four) full-time developers.
This enabled us, with the benefit of several years' worth of experience,
to redesign core parts of the library.

One example of redesigning core parts of the library is the rewrite of NumPy's
random number generation facilities.
These have been maintained but unchanged for many years to preserve bit-by-bit
compatibility of random number streams.  A rewrite that allowed for the
underlying pseudorandom algorithms to be exchanged and extended was completed
by a volunteer contributor and integrated by a paid developer.

Another ongoing example is a redesign of the data type system.  Data types
(dtypes for short) tell us more about the elements stored in a NumPy array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
datetimes, or Python objects, each identified by its own dtype.

%We are busy overhauling the datatype
%system to make it behave consistently and to simplify the creation of
%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
While the majority of users rely only on numerical dtypes,
there are a multitude of use cases for which the current implementation is unsuited, such as physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, and automatic
differentiation\cite{pyadolc}.
%Projects addressing these exist, each struggles with current
%limitations.  % TODO: I may need citations, we could link github issues.
The proposed overhaul of the datatype system will make it behave consistently and will simplify the creation of custom dtypes, both in C and in Python.

\section*{Array protocol ??}

We are increasingly focused on improving the library's interaction with external
packages.

@@

%"""
%NumPy’s high level ndarray API has been implemented several times outside of
%NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
%arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
%as various NumPy-like implementations in the deep learning frameworks, like
%TensorFlow and PyTorch.
%
%Similarly there are many projects that build on top of the NumPy API for
%labeled and indexed arrays (XArray), automatic differentiation (Autograd,
%Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
%etc. that add additional functionality on top of the NumPy API. Most of these
%project also implement a close variation of NumPy’s level high API.
%
%We would like to be able to use these libraries together, for example we would
%like to be able to place a CuPy array within XArray, or perform automatic
%differentiation on Dask array code. This would be easier to accomplish if code
%written for NumPy ndarrays could also be used by other NumPy-like projects.
%"""

@@

>>>>

@@

@@

@@

Data types (dtypes for short) tell us more about the elements stored in a NumPy
array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
datetimes, or Python objects, each identified by its own dtype.

@@

Sometimes, it is desirable to combine arrays of different dimensions.
One may, for example, want to add a scalar to a vector, or scale each column of
an array by a specific scalar value.
This is done through {\em broadcasting}, where one of the arrays is virtually
duplicated (i.e., without copying any data in memory), so that the shapes of
the operands match.
Broadcasting is illustrated in Figure~\ref{fig:broadcasting}.

@@

\section*{Changing computational landscape}

\fixme{Ralf will take a first pass on the section on 1/12.}

@@

As Python's popularity for scientific computing, data science, and machine
learning increased, many new use cases arose.

- GPUs

- distributed computing

- other languages w/ arrays

- massive explosion in data science, machine learning, and artificial intelligence

% NumPy also went from a small community of mostly expert users to
% one of the most widely used foundations of the scientific Python
% ecosystem.
% Now the user base is very different....

Another ongoing example is a redesign of the data type system.
While the majority of users rely only on numerical dtypes, there are a
multitude of use cases for which the current implementation is unsuited, such
as physical units\cite{astropy,Goldbaum2018,pint}, geometrical
objects\cite{pygeos}, and automatic differentiation\cite{pyadolc}.
The proposed overhaul of the datatype system will make it behave consistently
and will simplify the creation of custom dtypes, both in C and in Python.

\section*{Array protocols}

\fixme{NumPy’s high level ndarray API has been implemented several times outside of
NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
as various NumPy-like implementations in the deep learning frameworks, like
TensorFlow and PyTorch.}

\fixme{Similarly there are many projects that build on top of the NumPy API for
labeled and indexed arrays (XArray), automatic differentiation (Autograd,
Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
etc. that add additional functionality on top of the NumPy API. Most of these
project also implement a close variation of NumPy’s level high API.}

\fixme{We would like to be able to use these libraries together, for example we would
like to be able to place a CuPy array within XArray, or perform automatic
differentiation on Dask array code. This would be easier to accomplish if code
written for NumPy ndarrays could also be used by other NumPy-like projects.}

@@

@@

% aka why is this so successful
% some sense of ongoing work / future directions

Over the last decade, NumPy has attracted more scientist developers.
The project has strong culture of using software-engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy.
The project has comprehensive, high-quality documentation,
integrated with the source
code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
